---
title: "pca barplot"
output: pdf+document
layout: default
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.path='{{ site.url }}/images/pca-barplot-')
```

## Objective
I describe how to use R to perform a principal component analysis on a data frame and how to use the `ggplot2` package to plot the loadings on each PC axis with a bar graph.

## The Data
This example uses the `attitude` dataset that comes preloaded with R.

```{r}
head(attitude)
```

## PCA and Extracting PC Axes
We could be interested in relationships among the full dataset, or among a subset of variables. In both cases, use the base R function `prcomp` to perform a principal components analysis on the data. To perform a PCA on all columns in the dataset:

```{r}
attitude.pca <- prcomp(attitude)
```

The PC axes can be accessed with `attitude.pca$rotation`:

```{r}
PCaxes<-attitude.pca$rotation
PCaxes
```

Take the object with the principal components and turn it into a data frame. The variable names need to be moved from row names into their own column. I'm also going to add another fake categorical variable to illustrate how to include an additional 'layer' of information. The variable will be called `category` and I'll assign it a value of `positive` or `negative`.

```{r}
PCaxes<-as.data.frame(PCaxes)
PCaxes$variable<-rownames(PCaxes)
PCaxes$class <- c("positive","negative","positive","positive","positive","negative","positive")
```


## Reshaping the Data
Reshape the data frame so that it's in _long format_; each possible variable type (or combination of) has its own row with columns for variables and another for values of those variables. There are a few ways of doing this, including `melt` in the `reshape2` package and `gather` in the `tidyr` package. This example uses the latter. See more about `melt` [here](http://seananderson.ca/2013/10/19/reshape.html) and [here](http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/). 

```{r}
library(tidyr)

df<-tidyr::gather(data=PCaxes,key=axis,value=loading,1:7)
head(df)
```

## Plotting
The data is now structured so that plotting is convenient. Each unique combination of `variable`, `class`, and `PC1` is its own row. Let's try plotting this information using the package `ggplot2`. Generally, we want a bar graph that has a separate bar for each PC axis and variable, with the value of the bar representing the loading on that PC axis and variable. In this form, the graph will __not__ present data on the percent variance explained by each axis.

The basic pieces are the data frame (`df`) and adding a bar plot (via `geom_bar()`). If you're unfamiliar with ggplot, terms in `aes()` are used to define plotting and aesthetic elements based on the data frame. Here, the x-axis presents the factors in `variable` and the y-axis the numbers in `loading`. Calling `stat="identity"`, means that the heights of the bars represent values in the data; the default is `stat="bin"` where the heights represent counts of the number of cases in a group. `facet_wrap()` adds a panel for each value of the variable specified in the `facets` term. In this case, we want a unique panel for each PC axis. As you can see, this basically produces the kind of plot that we want. 

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.cap="Basic PCA barplot",fig.pos = 'p'}
library(ggplot2)

ggplot(data=df) +
  geom_bar(aes(x=variable,y=loading),stat="identity") +
  facet_wrap(facets=~axis,nrow=1)
```

That's the basic graph but it doesn't look very nice. There are a lot of aesthetic changes that can be made with the various ggplot2 functions. I give an example of what's possible below but other modifications are possible. 

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.cap="PCA barplot with groups identified by color",fig.pos = 'p'}
ggplot(df) +
  geom_bar(aes(x=variable,y=loading,fill=class),stat="identity") +
  scale_x_discrete(position="top") +
  scale_y_continuous(breaks = c(-.4,0,.4)) +
  coord_flip() +
  facet_wrap(~axis,nrow=1) +
  theme_bw()+
  theme(
    strip.background = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "left",
    axis.text.x = element_text(angle = 90, vjust = .5)
  )
```